---
title: Using Tidyr
date: '2017-02-20'
hidefromhome: true
---

I believe it's fitting that my first foray into blogging about `R` programming should be about a topic & package (`tidyr`) that are essential to a practicing Data Scientist's toolkit - data munging. Hadley Wickham's package [`tidyr`](https://github.com/hadley/tidyr) provides a number of convenience functions to, in his words, ensure that:

- Each column is a variable.
- Each row is an observation.

Why is this important you might ask? Well for a number of reasons but below are the most important two:

1. Many `R` functions expect a `data.frame` to be passed in as a parameter (i.e the fundamental data structure in R, which in most cases fits the definition of "tidy" above)
2. This standard, very akin to the relational database model developed by Codd, allows you to not have to reinvent a represtation at each go and rather focus on the analysis/exploration/visualization processes that got your excited about `R` programming in the first place.

Like most, I enjoy applied examples, and when I originally read Hadley's original [Rstudio post](https://blog.rstudio.org/2014/07/22/introducing-tidyr/) introducing `tidyr` I felt it was a bit lacking in explanation and didn't necessarily help `tidyr` to truly shine. The purpose of this post is to show you how `tidyr`'s main features compares to common SQL commands used to pivot/un-pivot data as well as parse values into columns (which occaisionally may be difficult -> impossible depending on your SQL flavor).

For the purpose of both clarity and brevity, I will only use the following packages in this post:

- `tidyr`: the package of interest. Specially we'll cover it's four main verbs:
    - `gather()` 
    - `spread()` 
    - `unite()`
    - `separate()` 
- `SQLDF`: a package developed developed by G. Grothendieck that allows one to write SQL against data.frames
- `dplyr`: another one of Hadely's packages that makes data manipulation in `R`, SQL-like
- `stringr`: yet another one of Hadley's packages for string manipulation
- `RMedicareData`: a personal package I developed which can be used to source example data from [data.medicare.gov/](https://data.medicare.gov/)

To demonstrate the functionality of `tidyr`, I will:

1. Source an example dataset
2. Demonstrate each of the `tidyr` functions of note
3. And finally, show the SQL, using `sqldf`, that would be required to conduct an analagous operation 

# Source our data
```{r, echo=TRUE, eval=FALSE}
library(RMedicareData)
list_data()
# source a dataset consisting of Readmissions and Death statistics as the hospital level
d <- get_data(59)
# take a glimpse
```

# Demo Spread
# Source our data
```{r, echo=TRUE, eval=FALSE}
library(tidyr)
# source a dataset consisting of Readmissions and Death statistics as the hospital level
d1 <- d %>% 
  select(Hospital.Name, Measure.Name, Score) %>% 
  spread(key = Measure.Name, value = Score)
# take a glimpse
```



